Checking Big Suffix and LCP Arrays by Probabilistic Methods

### 2017-03-20 draft-new/draft.tex
2017/03/20

{\color{blue} 
Abstract changed. \\
Many changes, not to mark. \\
Some marked in blue for your check. \\
The figures and tables should be near to the content talking about them. \\
Do not use too many: we, based, following, clearly...  \\
}

@@

Suffix and longest common prefix (LCP) arrays play an important role in various string processing tasks, such as data compression, pattern matching and genome assembly. In many applications, these two data structures make up the core part of a powerful full-text index, called enhanced suffix array~\cite{Abouelhodaa2004}, which is more space efficient than a suffix tree and applicable to emulating most searching functionalities provided by the latter in the same time complexity. The first algorithm for building suffix array~(SA) in internal memory was presented in~\cite{Manber1993}. From then on, much more effort has been put on designing efficient constructors for suffix array on different computation models~\cite{Karkkainen2003, Ko2003, Kim2003, Nong11, Dementiev2008, Ferragina2012, Manzini2004, Bingmann12, Karkkainen2014, Nong14, Nong15}. In respect of the research on LCP array construction algorithms, the existing works can be classified into two categories with regard to their input requirements, where the algorithms from the first category compute both suffix and LCP arrays at the same time with the original text only~\cite{Fischer11, Bingmann12, Flick2015}, and those from the second category carry out the computation by taking SA and/or Burrows-Wheeler transform (BWT) as additional inputs~\cite{Kasai2001,Karkkainen2009, Fischer11, Puglisi2008, Deo2013}. So far, the algorithms designed by the induced sorting~(IS) principle take linear time and space to run and get the best results on both internal and external memory~\cite{Nong11, esais}. In addition to the sequential algorithms, there are also parallel algorithms proposed to achieve high performance by fully using the available multi-core CPUs and/or GPUs~\cite{Osipov2012, Deo2013, Wang2015, Karkkainen2015, Karkkainen2016}.

@@

Our contribution comprises two methods to probabilistically verify any given suffix and LCP arrays. In principle, Method A checks the lexical order and the LCP-value of two neighboring suffixes in the suffix array by literally comparing the characters of their LCPs. For reducing the time complexity of a comparison between two sequences of characters, we use a Karp-Rabin fingerprinting technique to convert each sequence into a single integer, called fingerprint, and compare the fingerprints instead to check the equality of two sequences. The algorithm for Method A involves multiple scans and sorts on sets of $\mathcal{O}(n)$ fixed-size items. Its implementation in external memory suffers from a space bottleneck due to the large disk volume taken by each sort. 
To overcome this drawback, Method B first employs the fingerprinting technique to check a subset selected from the given suffix and LCP arrays, then it reuses the inducing process of the IS-based construction algorithm SA-IS \cite{Nong11} to produce the final suffix and LCP arrays from the verified subset and literally compares them with the input arrays to ensure the correctness of the latter. Our experiments indicate that the peak disk use of the programs for Algorithm~\ref{alg:2} designed by Method B is only about half as that of the program for Algorithm~\ref{alg:1} designed by Method A.

The remainder of this paper is organized as follows. Sections~\ref{sec:method1} and~\ref{sec:method2} describe the two methods and their algorithmic designs. Section~\ref{sec:experiment} conducts an experimental study for performance evaluation of our programs for these algorithms. Section~\ref{sec:conclusion} gives some concluding remarks.

@@

Notice that two equal substrings always share a common fingerprint, but the inverse is not true. It has been proven in~\cite{Karp1987} that the probability of a false match can be reduced to a negligible level by setting $L$ to a large value\footnote{This property is utilized in~\cite{Bille2013} to design a probabilistic algorithm for computing a sparse suffix array.}. Hence, we have:

@@

Fig.~\ref{fig:example} gives an illustrating example for utilizing Corollary~\ref{corollary:1} to check the input suffix and LCP arrays. Given that $L = 197$ and $\delta = 101$, lines 4-8 compute ${\sf fp}(0, p)$ iteratively according to Formulas~\ref{formula:1}-\ref{formula:2}. Then, lines 10-16 use these values to compute the fingerprints for all the target substrings. In more detail, consider the leftmost pair of neighboring suffixes in $sa$, i.e. ${\sf suf}(sa[0])$ and ${\sf suf}(sa[1])$, the substrings given by their LCP-value are $x[sa[0], sa[0] + lcp[1] - 1]$ and $x[sa[1], sa[1] + lcp[1] - 1]$, respectively. According to Formula~\ref{formula:3}, ${\sf fp}(sa[0], sa[0] + lcp[1] - 1)$ is equal to the difference between ${\sf fp}(0, sa[0] - 1)$ and ${\sf fp}(0, sa[0] + lcp[1] - 1)$, both have been calculated. Following the same way, ${\sf fp}(sa[1], sa[1] + lcp[1] - 1)$ is computed by reducing ${\sf fp}(0, sa[1] - 1)$ from ${\sf fp}(0, sa[1] + lcp[1] - 1)$. Hence, we obtain the fingerprints for these two substrings in lines 10-12 and see that they are equal to each other.

@@

		{\color{blue} sort the tuples in $ST_1$, $ST_2$ and $ST_3$ by the 1st components, respectively;} \label{alg:1:b}

@@

			\If{$ST_1.{\sf notEmpty}()$ {\sf and} $ST_1.{\sf top}().1st = i$}{~\label{alg:1:h}
				$e := ST_1.{\sf top}()$, $ST_1.{\sf pop}()$, $e.3rd := fp$, $ST_1'.{\sf push}(e)$
			}
			\Else{
				\Return false \hspace{5cm} // condition (1) is violated
			}~\label{alg:1:i}
			\While{$ST_2.{\sf notEmpty}()$ {\sf and} $ST_2.{\sf top}().1st = i$}{

@@

		{\color{blue} sort the tuples in $ST_1'$, $ST_2'$ and $ST_3'$ by the 2nd component, respectively;} \label{alg:1:e}

@@

				\Return false \hspace{5cm} // condition (2) or (3) is violated
			}	
		} \label{alg:1:g}

@@

Algorithm~\ref{alg:1} performs multiple scans and sorts on the arrays of $\mathcal{O}(n)$ fixed-size tuples in disks. Given RAM size $M$, disk size $D$ and block size $B$, all are in words, the time and I/O complexities for each scan are $\mathcal{O}(n)$ and $\mathcal{O}(n / B)$, respectively, while those for each sort are $\mathcal{O}(n\log_{M/ B}(n / B))$ and $\mathcal{O}((n / B)\log_{M / B}(n / B))$, respectively~\cite{Arge2013}. Algorithm~\ref{alg:1} reaches its peak disk use when sorting the tuples in lines~\ref{alg:1:b} and~\ref{alg:1:e}. Suppose the input string and the suffix/LCP array are encoded as $\alpha$- and $\beta$-byte integers, respectively, and each fingerprint is a $\gamma$-byte integer, it takes $(2 \cdot (2 \cdot \alpha + \beta))$ space for sorting the tuples of $ST_1/ST_1'$ and $(2 \cdot (2 \cdot \alpha + \beta + \gamma))$ for sorting the tuples of $ST_2/ST_2'$ and $ST_3/ST_3'$. For saving space, our program implementing the algorithm tackles $ST_1/ST_1'$, $ST_2/ST_2'$ and $ST_3/ST_3'$ separately and performs a single scan over $x$ for each of them to obtain the fingerprints, using less space but more time. The experiment in Section~\ref{sec:experiment} indicates that the peak disk use is 40 bytes per input character.

@@

We further to give another checking method using the induced sorting principle {\color{blue} \cite{sais,ka}}, , which requires much less space than Method A. For presentation convenience, we introduce some symbols and notations as below.

{\em Character and suffix classification.} All the characters in $x$ are classified into three types, namely L-, S- and S*-type. In detail, $x[i]$ is L-type if (1) $i = n - 1$ or (2) $x[i] > x[i + 1]$ or (3) $x[i] = x[i + 1]$ and $x[i + 1]$ is L-type; otherwise, $x[i]$ is S-type. Further, if $x[i]$ and $x[i + 1]$ are separately L-type and S-type, then $x[i + 1]$ is also S*-type. Moreover, a suffix is L-, S- or S*-type if its heading character is L-, S-, or S*-type, respectively.

{\em Suffix and LCP buckets.} All the suffixes in $sa$ are partitioned into multiple buckets and those of a common heading character are grouped into a single bucket that occupies a contiguous interval in $sa$. Each bucket can be further divided into two sub-buckets, where the left and the right parts  contain L- and S-type suffixes only, respectively. For short, we use ${\sf sa\_bkt}(c)$ to denote the bucket storing the suffixes starting with character $c$ and ${\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$ to denote its left/right sub-bucket. Accordingly, $lcp$ can be also split into multiple buckets, where ${\sf lcp\_bkt}(c)/{\sf lcp\_bkt_L}(c)/{\sf lcp\_bkt_S}(c)$ stores the LCP-values of suffixes in ${\sf sa\_bkt}(c)/{\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$, respectively.

{\em Suffix and LCP arrays for S*-type suffixes.} Given that the number of S*-type suffixes is $n_1$, $sa^*[0, n_1)$ stores all the S*-type suffixes arranged in lexical order, while $lcp^*[0] = 0$ and $lcp^*[i]$ records the LCP-value of ${\sf suf}(sa^*[i])$ and ${\sf suf}(sa^*[i - 1])$ for $i \in [1, n_1)$.

{\em Type Array.} The array $t$ records in $t[i]$ the type information of $x[i]$, where $t[i] = 1$ or $0$ if $x[i]$ is S- or L-type, respectively.

@@

The induced sorting principle has been extensively used to design efficient algorithms for constructing the suffix and LCP arrays in internal or external memory {\color{blue} \cite{sais,dsais,emsads,esais,inducelcp}}. Such a construction algorithm mainly consists of a reduction phase for computing $sa^*$ and $lcp^*$, followed by an induction phase for inducing $sa$ and $lcp$ from $sa^*$ and $lcp^*$~\footnote{An overview of the induction phase is given in Appendix~\ref{sec:appendix}.}. Given that $sa^*$ and $lcp^*$ are already known, we can induce the final suffix and LCP arrays from them. This suggests a checking method based on Lemma~\ref{lemma:2}.

@@

We further to design Algorithm~\ref{alg:2} for checking the conditions of Corollary~\ref{corollary:2}. The first step is to compute and verify $sa^*$ and $lcp^*$. Similar to Method A, the fingerprinting technique is employed to probabilistically check the correctness of $sa^*$ and $lcp^*$.  The array $sa^*$ can be produced by sequentially retrieving the S*-type suffixes from $sa$ and the LCP-value of two successive S*-type suffixes in $sa$, say ${\sf suf}(sa[i])$ and ${\sf suf}(sa[j])$, is equal to the minimal of $\{lcp[i + 1], ..., lcp[j - 1], lcp[j]\}$. The algorithm first sorts all the suffixes in $sa$ by their starting positions~(lines~\ref{alg:2:a}-\ref{alg:2:b}) and then scans $x$ once to get the S*-type suffixes~(lines~\ref{alg:2:c}-\ref{alg:2:d}). After that, it puts these S*-type suffixes back in their lexical order and outputs them one by one to generate $sa^*$~(lines~\ref{alg:2:g}-\ref{alg:2:h}). Meanwhile, it calculates the LCP-value for each pair of the neighboring suffixes in $sa^*$ by tracing the minimum in the $lcp$ interval between these two suffixes. Notice that, we check condition 1 when visiting the suffixes in their position order (lines~\ref{alg:2:e}-\ref{alg:2:f}), and check conditions 2 and 3 by Algorithm~\ref{alg:1} with $sa^*$ and $lcp^*$ as input. Suppose $sa^*$ and $lcp^*$ are both correct, then Algorithm~\ref{alg:2} invokes the inducing process of an existing IS-based construction algorithm to induce $sa'$ and $lcp'$ from the two verified arrays $sa^*$ and $lcp^*$~(line~\ref{alg:2:k}) and compares them with $sa$ and $lcp$ to complete the whole checking process~(lines~\ref{alg:2:l}-\ref{alg:2:m}).

@@

				\Return false  \hspace{1cm} // condition (1) is violated
			}

@@

		\Return false \hspace{1cm} // conditions (2) or (3) is violated
	}\label{alg:2:j}

@@

			\Return false \hspace{1cm}	// condition (4) is violated
		}	
	}\label{alg:2:m}

@@

{\bf Assume that the alphabet $\Sigma$ is of size $\mathcal{O}(1)$, we can check $sa$ and $lcp$ without storing the induced suffix and LCP arrays in Algorithm~\ref{alg:2}.} The idea is to compare the induced suffix/LCP items with their corresponding items in $sa/lcp$ during the inducing process. Specifically, when a suffix/LCP item $v_1$ is induced into a bucket, we check if it is equal to the corresponding item $v_2$ in $sa/lcp$. If $v_1 = v_2$, then $v_2$ is correct and we further use this value to induce the remaining suffix/LCP items. The key point is to quickly retrieve the items of $sa/lcp$ in external memory. This can be done by conducting sequential I/O operations if we provide a read pointer together with a buffer for each suffix/LCP sub-bucket. We describe below more details of the modified inducing process, where $lp_1/lp_2$ and $sp_1/sp_2$ are file pointer arrays for reading items from the sub-buckets of $sa$ and $lcp$.

@@

Algorithm~\ref{alg:2} mainly consists of two parts, where the first part for checking $sa^*$ and $lcp^*$ can be done within sorting complexity and the second part for checking $sa$ and $lcp$ can be done in linear time. In our current program for this algorithm, the peak disk use is reached in part two, specifically, when computing the BWT from $sa$ and $x$ for use in the inducing process.

@@

The above observations indicate that ProgB reaches its peak disk use when checking the final suffix and LCP arrays during the inducing process, i.e., the inducing process constitutes the performance bottleneck of the whole algorithm. By adopting the space optimization scheme introduced in Section~\ref{sec:method2:algorithm}, we adapt Algorithm~\ref{alg:2} and evaluate the tuned version of ProgB, called ProgB+, in comparison with ProgA and ProgB. Fig.~\ref{fig:performance_analysis} shows that the maximum space requirement for ProgB+ is about $21n$, which is much less than that of ProgB and even only half as that of ProgA. In addition, progB+ outperforms its prototype with respect to time and I/O efficiency and is faster than ProgA when handling "proteins". We also investigate the performance trend of the three programs on the prefix of "enwiki" with the length varying in $\{1, 2, 4, 8\}$ GiB. In Figure~\ref{fig:performance_analysis2}, the peak disk use for each program remains unchanged, but their speed become slower as the prefix length increases due to the performance degradation of the external memory sorter used in our programs. This can be also observed from Table~\ref{tbl:2}, where ProgB+ keeps the I/O volume around $90n$ with the prefix length of "enwiki" varying from $1$ to $8$ GiB but its running time rises from $1.05$ to $1.33$.

@@

We select these programs because they are currently the fastest suffix and LCP arrays builders available to us. A runtime breakdown of the programs for these solutions on the prefixes of "enwiki" is given in Table~\ref{tbl:3}. The program for Solution 2 is about two times faster than that for Solution 1 and twice as fast as ProgA, which is mainly due to the high speed of pSAscan in this experiment. However, it is worthy of pointing out that both pSAscan and Sparse-$\phi$ are of the time and I/O complexities proportional to $n^2/M$. This is much higher than eSAIS and our checking algorithms when $n$ increases, and thus poses a strict limitation to the scalability of Solution 2. As reported in~\cite{Karkkainen2015}, when $n$ is considerably greater than $M$, eSAIS is much more time and I/O efficient than pSAscan. In this experiment, pSAscan builds the SA for "enwiki\_8g" in time double as that for "enwiki\_1g". {\bf For big $n$, it is more reasonable to compare the results of our programs with that of Solution 1.}

@@

It should be noticed that our programs are coded for experimental study only. From engineering aspects, there is still a big margin for better implementation. For example, several algorithms for induced sorting a suffix and/or LCP arrays were proposed these years \cite{Nong15,Nong14,Bingmann12}, with different methods for solving the key problem of retrieving the preceding character of a sorted suffix in the inducing process. A recent work \cite{Karkkainen2017} engineering these induced sorting methods with some implementation optimizing techniques achieves a significant improvement over the previous results. As reported, the peak disk use is around $8n$ for 40-bit integers. Because the induced sorting process is the performance bottleneck for Method B, it is reasonable to expect that a better engineering implementation of the method will yield a remarkable improvement in both time and space. {\bf An optimized engineering of our methods is out of the scope of this paper and will be addressed elsewhere.}

@@

In this paper, we propose two methods for probabilistically checking the given suffix and LCP arrays. Theoretically, the external-memory algorithms designed by these methods have better time and I/O complexities compared to the existing fastest construction algorithms. Our experimental results indicate that the current programs for Algorithm~\ref{alg:2} designed by Method B run slower than that for Algorithm~\ref{alg:1} designed by Method A, but they are much more space-efficient than the latter. As discussed in Section~\ref{sec:experiment}, there still remains much room for improving the implementations of the proposed algorithms. Our experimental programs can be further optimized to achieve higher performance, in particular for checking arrays of constant alphabets that are most common in practice. The optimized programs will run much faster and use at most $n$ integers as the working space in addition to the input arrays.

@@

\appendices
\section{Overview on the Induction Phase} \label{sec:appendix}

>>>>

@@

Suffix and longest common prefix (LCP) arrays play an important role in various string processing tasks, such as data compression, pattern matching and genome assembly. In many applications, these two data structures make up the core part of a powerful full-text index, called enhanced suffix array~\cite{Abouelhodaa2004}, which is more space efficient than a suffix tree and applicable to emulating most searching functionalities provided by the latter in the same time complexity. The first algorithm for building suffix array~(SA) in internal memory was presented in~\cite{Manber1993}. From then on, much more effort has been put on designing efficient constructors for suffix array on different computation models~\cite{Karkkainen2003, Ko2003, Kim2003, Nong11, Dementiev2008, Ferragina2012, Manzini2004, Bingmann12, Karkkainen2014, Nong14, Nong15}. In respect of the research on LCP array construction algorithms, the existing works can be classified into two categories with regard to their input requirements, where the algorithms from the first category compute both suffix and LCP arrays at the same time with the original text only~\cite{Fischer11, Bingmann12, Flick2015}, and those from the second category carry out the computation by taking SA and/or Burrows-Wheeler transform (BWT) as additional inputs~\cite{Kasai2001,Karkkainen2009, Fischer11, Puglisi2008, Deo2013}. So far, the algorithms designed by the induced sorting~(IS) principle take linear time and space to run and get the best results on both internal and external memory~\cite{Nong11, Bingmann12}. In addition to the sequential algorithms, there are also parallel algorithms proposed to achieve high performance by fully using the available multi-core CPUs and/or GPUs~\cite{Osipov2012, Deo2013, Wang2015, Karkkainen2015, Karkkainen2016}.

@@

Our contribution comprises two methods to probabilistically verify any given suffix and LCP arrays. In principle, Method A checks the lexical order and the LCP-value of two neighboring suffixes in the suffix array by literally comparing the characters of their LCPs. For reducing the time complexity of a comparison between two sequences of characters, we use a Karp-Rabin fingerprinting technique to convert each sequence into a single integer, called fingerprint, and compare the fingerprints instead to check the equality of two sequences. The algorithm for Method A involves multiple scans and sorts on sets of $\mathcal{O}(n)$ fixed-size items. Its implementation in external memory suffers from a space bottleneck due to the large disk volume taken by each sort. 
To overcome this drawback, Method B first employs the fingerprinting technique to check a subset selected from the given suffix and LCP arrays, 
%then it reuses the inducing process of the IS-based construction algorithm SA-IS \cite{Nong11} to produce the final suffix and LCP arrays from the verified subset and literally compares them with the input arrays to ensure the correctness of the latter. Our experiments indicate that the peak disk use of the programs for Algorithm~\ref{alg:2} designed by Method B is only about half as that of the program for Algorithm~\ref{alg:1} designed by Method A. 
{\color{red}{then it utilizes the IS method to produce the final suffix and LCP arrays from the verified subset and literally compares them with the input arrays to ensure the correctness of the latter. Our experiments indicate that the program for Algorithm~\ref{alg:2} designed by Method B only takes around half as much disk space as the program for Algorithm~\ref{alg:1} designed by Method A.}}

The remainder of this paper is organized as follows. Sections~\ref{sec:method1} and~\ref{sec:method2} describe the two methods and their algorithmic designs. Section~\ref{sec:experiment} conducts an experimental study for performance evaluation of our programs for these algorithms. {\color{red}{Finally}}, Section~\ref{sec:conclusion} gives some concluding remarks.

@@

Notice that two equal substrings always share a common fingerprint, but the inverse is not true. It has been %proven
{\color{red}{proved}} in~\cite{Karp1987} that the probability of a false match can be reduced to a negligible level by setting $L$ to a large value\footnote{This property is utilized in~\cite{Bille2013} to design a probabilistic algorithm for computing a sparse suffix array.}. Hence, we have:

@@

Fig.~\ref{fig:example} gives an illustrating example for utilizing Corollary~\ref{corollary:1} to check the input suffix and LCP arrays. Given that $L = 197$ and $\delta = 101$, lines 4-8 compute ${\sf fp}(0, p)$ iteratively according to Formulas~\ref{formula:1}-\ref{formula:2}. Then, {\color{red}{lines 10-20}} use these values to compute the fingerprints for all the target substrings. In more detail, consider the leftmost pair of neighboring suffixes in $sa$, i.e. ${\sf suf}(sa[0])$ and ${\sf suf}(sa[1])$, the substrings given by their LCP-value are $x[sa[0], sa[0] + lcp[1] - 1]$ and $x[sa[1], sa[1] + lcp[1] - 1]$, respectively. According to Formula~\ref{formula:3}, ${\sf fp}(sa[0], sa[0] + lcp[1] - 1)$ is equal to the difference between ${\sf fp}(0, sa[0] - 1)$ and ${\sf fp}(0, sa[0] + lcp[1] - 1)$, both have been calculated. Following the same way, ${\sf fp}(sa[1], sa[1] + lcp[1] - 1)$ is computed by reducing ${\sf fp}(0, sa[1] - 1)$ from ${\sf fp}(0, sa[1] + lcp[1] - 1)$. Hence, we obtain the fingerprints for these two substrings in {\color{red}{lines 10-14}} and see that they are equal to each other.

@@

		sort the tuples in $ST_1$, $ST_2$ and $ST_3$ by the 1st components, respectively; \label{alg:1:b}

@@

			\If{$ST_1.{\sf notEmpty}()$ {\sf and} $ST_1.{\sf top}().1st = i$}{~\label{alg:1:h}
				$e := ST_1.{\sf top}()$, $ST_1.{\sf pop}()$, $e.3rd := fp$, $ST_1'.{\sf push}(e)$
			}
			\Else{
				\Return false \hspace{5cm} // condition 1 is violated
			}~\label{alg:1:i}
			\While{$ST_2.{\sf notEmpty}()$ {\sf and} $ST_2.{\sf top}().1st = i$}{

@@

		sort the tuples in $ST_1'$, $ST_2'$ and $ST_3'$ by the 2nd component, respectively; \label{alg:1:e}

@@

				\Return false \hspace{5cm} // condition 2 or 3 is violated
			}	
		} \label{alg:1:g}

@@

Algorithm~\ref{alg:1} performs multiple scans and sorts on the arrays of $\mathcal{O}(n)$ fixed-size tuples in disks. Given RAM size $M$, disk size $D$ and block size $B$, all are in words, the time and I/O complexities for each scan are $\mathcal{O}(n)$ and $\mathcal{O}(n / B)$, respectively, while those for each sort are $\mathcal{O}(n\log_{M/ B}(n / B))$ and $\mathcal{O}((n / B)\log_{M / B}(n / B))$, respectively~\cite{Arge2013}. Algorithm~\ref{alg:1} reaches its peak disk use when sorting the tuples in lines~\ref{alg:1:b} and~\ref{alg:1:e}. Suppose the input string and the suffix/LCP array are encoded as $\alpha$- and $\beta$-byte integers, respectively, and each fingerprint is a $\gamma$-byte integer, it takes $(2 \cdot (\alpha +2 \cdot \beta))$ space for sorting the tuples of $ST_1/ST_1'$ and $(2 \cdot (\alpha + 2 \cdot \beta + \gamma))$ for sorting the tuples of $ST_2/ST_2'$ and $ST_3/ST_3'$. For saving space, our program implementing the algorithm tackles $ST_1/ST_1'$, $ST_2/ST_2'$ and $ST_3/ST_3'$ separately and performs a single scan over $x$ for each of them to obtain the fingerprints, using less space but more time. The experiment in Section~\ref{sec:experiment} indicates that {\color{red}{the disk use is 40 times the size of $x$.}}

@@

We further to give another checking method using the induced sorting principle \cite{Nong11,Dementiev08}, which requires much less space than Method A. For presentation convenience, we introduce some symbols and notations as below.

Character and suffix classification. All the characters in $x$ are classified into three types, namely L-, S- and S*-type. In detail, $x[i]$ is L-type if (1) $i = n - 1$ or (2) $x[i] > x[i + 1]$ or (3) $x[i] = x[i + 1]$ and $x[i + 1]$ is L-type; otherwise, $x[i]$ is S-type. Further, if $x[i]$ and $x[i + 1]$ are separately L-type and S-type, then $x[i + 1]$ is also S*-type. Moreover, a suffix is L-, S- or S*-type if its heading character is L-, S-, or S*-type, respectively.

Suffix and LCP buckets. All the suffixes in $sa$ are partitioned into multiple buckets and those of a common heading character are grouped into a single bucket that occupies a contiguous interval in $sa$. Each bucket can be further divided into two sub-buckets, where the left and the right parts  contain L- and S-type suffixes only, respectively. For short, we use ${\sf sa\_bkt}(c)$ to denote the bucket storing the suffixes starting with character $c$ and ${\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$ to denote its left/right sub-bucket. Accordingly, $lcp$ can be also split into multiple buckets, where ${\sf lcp\_bkt}(c)/{\sf lcp\_bkt_L}(c)/{\sf lcp\_bkt_S}(c)$ stores the LCP-values of suffixes in ${\sf sa\_bkt}(c)/{\sf sa\_bkt_L}(c)/{\sf sa\_bkt_S}(c)$, respectively.

Suffix and LCP arrays for S*-type suffixes. Given that the number of S*-type suffixes is $n_1$, $sa^*[0, n_1)$ stores all the S*-type suffixes arranged in lexical order, while $lcp^*[0] = 0$ and $lcp^*[i]$ records the LCP-value of ${\sf suf}(sa^*[i])$ and ${\sf suf}(sa^*[i - 1])$ for $i \in [1, n_1)$.

Type Array. The array $t$ records in $t[i]$ the type information of $x[i]$, where $t[i] = 1$ or $0$ if $x[i]$ is S- or L-type, respectively.

@@

The induced sorting principle has been extensively used to design efficient algorithms for constructing the suffix and LCP arrays in internal or external memory \cite{Nong11,Nong14, Nong15, Bingmann12,Fischer11}. Such a construction algorithm mainly consists of a reduction phase for computing $sa^*$ and $lcp^*$, followed by an induction phase for inducing $sa$ and $lcp$ from $sa^*$ and $lcp^*$~\footnote{An overview of the induction phase is given in Appendix~\ref{sec:appendix}.}. Given that $sa^*$ and $lcp^*$ are already known, we can induce the final suffix and LCP arrays from them. This suggests a checking method based on Lemma~\ref{lemma:2}.

@@

We further to design Algorithm~\ref{alg:2} for checking the conditions of Corollary~\ref{corollary:2}. The first step is to compute and verify $sa^*$ and $lcp^*$. Similar to Method A, the fingerprinting technique is employed to probabilistically check the correctness of $sa^*$ and $lcp^*$.  The array $sa^*$ can be produced by sequentially retrieving the S*-type suffixes from $sa$ and the LCP-value of two successive S*-type suffixes in $sa$, say ${\sf suf}(sa[i])$ and ${\sf suf}(sa[j])$, is equal to the minimal of $\{lcp[i + 1], ..., lcp[j - 1], lcp[j]\}$. The algorithm first sorts all the suffixes in $sa$ by their starting positions~(lines~\ref{alg:2:a}-\ref{alg:2:b}) and then scans $x$ once to get the S*-type suffixes~(lines~\ref{alg:2:c}-\ref{alg:2:d}). After that, it puts these S*-type suffixes back in their lexical order and outputs them one by one to generate $sa^*$~(lines~\ref{alg:2:g}-\ref{alg:2:h}). Meanwhile, it calculates the LCP-value for each pair of the neighboring suffixes in $sa^*$ by tracing the minimum in the $lcp$ interval between these two suffixes. Notice that, we check condition 1 when visiting the suffixes in their position order (lines~\ref{alg:2:e}-\ref{alg:2:f}), and check conditions 2 and 3 by Algorithm~\ref{alg:1} with $sa^*$ and $lcp^*$ as input. Suppose $sa^*$ and $lcp^*$ are both correct, %then Algorithm~\ref{alg:2} invokes the inducing process of an existing IS-based construction algorithm
{\color{red}{then Algorithm~\ref{alg:2} invokes an inducing process for employing the IS method}} to induce $sa'$ and $lcp'$ from the two verified arrays $sa^*$ and $lcp^*$~(line~\ref{alg:2:k}) and compares them with $sa$ and $lcp$ to complete the whole checking process~(lines~\ref{alg:2:l}-\ref{alg:2:m}).

@@

				\Return false  \hspace{1cm} // condition 1 is violated
			}

@@

		\Return false \hspace{1cm} // conditions 2 or 3 is violated
	}\label{alg:2:j}

@@

			\Return false \hspace{1cm}	// condition 4 is violated
		}	
	}\label{alg:2:m}

@@

Assume that the alphabet $\Sigma$ is of size $\mathcal{O}(1)$, we can check $sa$ and $lcp$ without storing the induced suffix and LCP arrays in Algorithm~\ref{alg:2}. The idea is to compare the induced suffix/LCP items with their corresponding items in $sa/lcp$ during the inducing process. Specifically, when a suffix/LCP item $v_1$ is induced into a bucket, we check if it is equal to the corresponding item $v_2$ in $sa/lcp$. If $v_1 = v_2$, then $v_2$ is correct and we further use this value to induce the remaining suffix/LCP items. The key point is to quickly retrieve the items of $sa/lcp$ in external memory. This can be done by conducting sequential I/O operations if we provide a read pointer together with a buffer for each suffix/LCP sub-bucket. We describe below more details of the modified inducing process, %where $lp_1/lp_2$ and $sp_1/sp_2$ are file pointer arrays for reading items from the sub-buckets of $sa$ and $lcp$.
{\color{red}{where $lp_1/lp_2$ and $sp_1/sp_2$ indicate the next items to be visited in the L-type and S-type sub-buckets, respectively.}}

@@

{\color{red}{Algorithm~\ref{alg:2} mainly consists of two steps, where the first step for checking $sa^*$ and $lcp^*$ can be done within sorting complexity and the second step for checking $sa$ and $lcp$ can also be done in sorting complexity for using external-memory sorters and priority queues. In our current program for this algorithm, the peak disk use is reached in step two, specifically, when computing the BWT from $sa$ and $x$ for use in the inducing process. }}

@@

The above observations indicate that ProgB reaches its peak disk use when checking the final suffix and LCP arrays during the inducing process, i.e., the inducing process constitutes the performance bottleneck of the whole algorithm. By adopting the space optimization scheme introduced in Section~\ref{sec:method2:algorithm}, we adapt Algorithm~\ref{alg:2} and evaluate the tuned version of ProgB, called ProgB+, in comparison with ProgA and ProgB. Fig.~\ref{fig:performance_analysis} shows that the maximum space requirement for ProgB+ is about $21n$, which is much less than that of ProgB and even only half as that of ProgA. In addition, progB+ outperforms its prototype with respect to time and I/O efficiency and is faster than ProgA when handling "uniprot". We also investigate the performance trend of the three programs on the prefix of "enwiki" with the length varying in $\{1, 2, 4, 8\}$ GiB. In Figure~\ref{fig:performance_analysis2}, the peak disk use for each program remains unchanged, but their speed become slower as the prefix length increases due to the performance degradation of the external memory sorter used in our programs. This can be also observed from Table~\ref{tbl:2}, where ProgB+ keeps the I/O volume around $90n$ with the prefix length of "enwiki" varying from $1$ to $8$ GiB but its running time rises from $1.05$ to $1.33$.

@@

We select these programs because they are currently the fastest suffix and LCP arrays builders available to us. A runtime breakdown of the programs for these solutions on the prefixes of "enwiki" is given in Table~\ref{tbl:3}. The program for Solution 2 is about two times faster than that for Solution 1 and twice as fast as ProgA, which is mainly due to the high speed of pSAscan in this experiment. However, it is worthy of pointing out that both pSAscan and Sparse-$\phi$ are of the time and I/O complexities proportional to $n^2/M$. This is much higher than eSAIS and our checking algorithms when $n$ increases, and thus poses a strict limitation to the scalability of Solution 2. As reported in~\cite{Karkkainen2015}, when $n$ is considerably greater than $M$, eSAIS is much more time and I/O efficient than pSAscan. In this experiment, pSAscan builds the SA for "enwiki\_8g" in time double as that for "enwiki\_1g". For big $n$, it is more reasonable to compare the results of our programs with that of Solution 1.

@@

It should be noticed that our programs are coded for experimental study only. From engineering aspects, there is still a big margin for better implementation. For example, several algorithms for induced sorting a suffix and/or LCP arrays were proposed these years \cite{Nong15,Nong14,Bingmann12}, with different methods for solving the key problem of retrieving the preceding character of a sorted suffix in the inducing process. A recent work \cite{Karkkainen2017} engineering these induced sorting methods with some implementation optimizing techniques achieves a significant improvement over the previous results. As reported, the peak disk use is around $8n$ for 40-bit integers. Because the induced sorting process is the performance bottleneck for Method B, it is reasonable to expect that a better engineering implementation of the method will yield a remarkable improvement in both time and space. An optimized engineering of our methods is out of the scope of this paper and will be addressed elsewhere.

@@

In this paper, we propose two methods for probabilistically checking the given suffix and LCP arrays. Theoretically, the external-memory algorithms designed by these methods have better time and I/O complexities compared to the existing fastest construction algorithms. Our experimental results indicate that the current programs for Algorithm~\ref{alg:2} designed by Method B run slower than that for Algorithm~\ref{alg:1} designed by Method A, but they are much more space-efficient than the latter. As discussed in Section~\ref{sec:experiment}, there still remains much room for improving the implementations of the proposed algorithms. Our experimental programs can be further optimized to achieve higher performance, in particular for checking arrays of constant alphabets that are most common in practice. {\color{red}{The optimized programs will run much faster and use at most $n$ integers as the working space in addition to the input arrays.}  \color{blue}{Should be deleted. This may not be correct, because it takes more space to verify $sa_L/sa_S$ and $lcp_L/lcp_S$ by Algorithm~\ref{alg:1}.}}

@@

\newpage
\appendices
\section{Overview on the Induction Phase} \label{sec:appendix}
